A comparison of model-based and design-based approaches for spatial data.

### 2021-06-08 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
Added citations and started introductory paragraph.

@@

de1990model: early comparison of design v model in spatial context

Several comparisons between modegitl-based and design-based for spatial data have been considered, but they tend to compare spatial models to IRS (Independent Random Sampling) or SIRS (stratified IRS). No comparison has been made between spatial models and spatially balanced design-based approaches.

@@

* brus1997random - design and model based explanation and simulation study for spatial means, make a decision tree for choosing among approaches

@@

The design-based approach to inference uses characteristics of the sampling design to estimate parameters of interest. The design-based approach assumes the data, $\mathbf{Y}$, are fixed. Randomness is incorporated through the random selection of sampling units based on probabilities derived using the sampling design itself. Some examples of commonly used sampling designs include independent random sampling (IRS), stratified random sampling, and cluster sampling. Viewing the data as fixed and incorporating randomness through the sampling design yields estimators having very few assumptions. Asymptotic properties of these types of estimators are derived using limiting arguments. Means and totals, for example, are asymptotically Gaussian. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

@@

Other model based stuff:
chan2020bayesian: Bayesian model-based approach.
Based on splines: 
Cicchitelli, G. and Montanari, G. E. (2012). “Model-Assisted Estimation of a Spatial Population
Mean.” International Statistical Review, 80(1): 111–126.
<!-- \citep{Cressie2015statistics, Schabenberger2017statistical} -->

@@

@@

Several comparisons between the two paradigms have been made in many different contexts. @sterba2009alternative give some history of the two paradigms as well as some applications in psychological contexts. @cooper2006sampling review the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. We note that while there has been substantial research and development into estimators that use both design and model-based principles (see e.g. CITE some more f these), our goal is not to expand upon or improve these methods.

@@

The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population totals (and means). If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$, denoted by $\hat{\tau}_{HT}$, is given by
\begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Y_i \pi_i^{-1},
\end{align}
where $Y_i$ is the observed value of the $i$th sampling unit, and $\pi_i$ is the inclusion probability of the $i$th sampling unit. Estimating the variance of $\hat{\tau}_{HT}$ directly relies on knowing the probability that $s_i$ and $s_j$ are both included in the sample (second-order inclusion probability), which can be unfeasible to compute. Furthermore, this variance estimator does not incorporate spatial locations. To address these challenges, [@stevens2003variance] proposed an alternative variance estimator: the local neighborhood variance estimator. The local neighborhood variance estimator does not require second-order inclusion probabilities. It also incorporates spatial locations, which tends to reduce the estimated variance of compared $\hat{\tau}_{HT}$ to a variance estimator ignoring spatial location [@stevens2003variance].

>>>>

@de1990model give an early comparison of the paradigms for spatial data, quashing the belief that classical sampling-based methods could not be used for spatially correlated data. Thereafter, several comparisons between model-based and design-based for spatial data have been considered, but they tend to compare spatial models to IRS (Independent Random Sampling) or SIRS (stratified IRS) [@brus1997random]. No comparison has been made between spatial models and spatially balanced design-based approaches.

@@

@@

@@

The design-based approach to inference uses characteristics of the sampling design to estimate parameters of interest. The design-based approach assumes the data, $\mathbf{Z}$, are fixed. Randomness is incorporated through the random selection of sampling units based on probabilities derived using the sampling design itself. Some examples of commonly used sampling designs include independent random sampling (IRS), stratified random sampling, and cluster sampling. Viewing the data as fixed and incorporating randomness through the sampling design yields estimators having very few assumptions. Asymptotic properties of these types of estimators are derived using limiting arguments. Means and totals, for example, are asymptotically Gaussian. @sarndal2003model and @lohr2009sampling provide thorough reviews of the design-based approach.

@@

<!-- \citep{Cressie2015statistics, Schabenberger2017statistical} -->

@@

We only use FPBK in this paper in order to focus more on comparing the design-based and model-based approaches. However, k-nearest-neighbors [@fix1951discriminatory; @ver2013comparison], random forest [@breiman2001random], Bayesian models [@chan2020bayesian], among others, can also be used to obtain predictions for a mean or total from spatially correlated responses in a finite population setting.

@@

Several comparisons between the two paradigms have been made in many different contexts. @sterba2009alternative give some history of the two paradigms as well as some applications in psychological contexts. @cooper2006sampling review the two approaches in an ecological context before introducing a "model-assisted" variance estimator that combines aspects from each approach. We note that while there has been substantial research and development into estimators that use both design and model-based principles (see e.g. @cicchitelli2012model, @chan2020bayesian for a Bayesian approach, CITE some more of these), our goal is not to expand upon or improve these methods.

@@

The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population totals (and means). If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$, denoted by $\hat{\tau}_{HT}$, is given by
\begin{align}\label{eq:htest}
  \hat{\tau}_{HT} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ is the observed value of the $i$th sampling unit, and $\pi_i$ is the inclusion probability of the $i$th sampling unit. Estimating the variance of $\hat{\tau}_{HT}$ directly relies on knowing the probability that $s_i$ and $s_j$ are both included in the sample (second-order inclusion probability), which can be unfeasible to compute. Furthermore, this variance estimator does not incorporate spatial locations. To address these challenges, [@stevens2003variance] proposed an alternative variance estimator: the local neighborhood variance estimator. The local neighborhood variance estimator does not require second-order inclusion probabilities. It also incorporates spatial locations, which tends to reduce the estimated variance of compared $\hat{\tau}_{HT}$ to a variance estimator ignoring spatial location [@stevens2003variance].
